Working from home and digital divides: resilience in the time of a pandemic

### 2020-12-30 paper/v1/broadband.speed.covid.tex
resizing images

\begin{abstract}
This paper offers a new perspective on telecommuting from the viewpoint
of the complex web of digital divides. Using the UK as a case study,
this paper studies how the quality and reliability of internet service,
as reflected in \emph{experienced} vis-à-vis \emph{advertised} upload
internet speeds during the spring 2020 lockdown, may reinforce or
redress the spatial and social dimensions of digital divisions. Fast,
reliable internet connections are necessary for the population to be
able to work from home. Although not every place hosts individuals in
occupations which allow for telecommuting nor with the necessary skills
to effectively use the internet to telecommute, we raise the issue that
without good internet connectivity, these places would struggle more to
achieve economic resilience in a period like the current pandemic. We
employ data on individual broadband speed tests and state-of-the-art
time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet
speeds. We then associate these clusters of local authorities with their
socioeconomic and geographic characteristics to explore how they overlap
with or diverge from the existing economic and digital geography of the
UK. Our analysis enables us to better understand how the spatial and
social distribution of both occupations and online accessibility
intersect to enable or hinder the practice of telecommuting at a time of
extreme demand.
\end{abstract}

@@

In this analysis, the terms `telecommuting' and `working from
home\footnote{See also the popular social media hashtag \#WFM}' are used
interchangeably, as most remote labour during the Covid-19 crisis was
carried out in the homes of individual employees rather than any other
location \citep{eurofound2020}. However, it should be noted that
previous research has explored how telecommuting can occur in other
places, including satellite offices or on public transport
\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
also used a variety of definitions to measure the level of telecommuting
within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or
part-time, and those who use digital technologies to work remotely
full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. For example, studies from the United States, the
Netherlands, and the UK found that telecommuters are most likely to hold
professional, managerial, and technical occupations where the workforce
is better educated and wealthier, and that there is suppressed demand
among women and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.

Opportunities for working from home during the current pandemic have
likewise not been equally spread across the workforce.
\citet{NBERw26948} indicated that in the US, managers, educators, as
well as those working in computer-related occupations, finance, and law
can easily work from home, and that occupations with opportunities to
telecommute are associated with higher earnings. This is not the case
for the workforce occupied in more spatially fixed occupations, from
farming, construction and manufacturing to hospitality and care
services. In the US, these occupations tend to be lower-income,
non-white, without a university degree, live in rental accommodation and
lack health insurance \citep{NBERw27085}. Similar trends can be observed
for other countries. For example, \(75\)\% of workers with tertiary
education worked from home in Europe during spring \(2020\), whilst only
\(34\)\% of workers with secondary education and \(14\)\% of those
primary education did so \citep{eurofound2020}.

@@

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
These returns have been even greater during the Covid-19 crisis, when
those with the capability to telecommute also have the ability to
maintain their employment whilst protecting their health. However, the
success of these arrangements has been dependent upon the first level
digital divide, which is associated with access to and quality of
internet connectivity. \citet{SALEMINK2017360} provides a systematic
review of the pre-pandemic, first level digital divide in infrastructure
quality between urban and rural areas in various advanced economies.
Rural areas, predictably, fare worse. Yet whether this variation in
infrastructure quality affects the spatial footprint of telecommuting
has not previously been discussed, in part because telecommuting has not
previously been the cause of greatest demand and pressure on internet
services.

There are indications that those who purchase high speed connections
consume more data of all sorts and use their connections for a variety
of purposes \citep{hauge2011consumer}. There is also a correlation
between access to internet services and a reduction in household
transport spend \citep{bris2017ict}. Independently of whether the
additional internet use and reduced travel is because of increased
telecommuting, these studies suggest that better internet services
enable households to make savings and efficiencies, an example of the
first level digital divide reinforcing the third level.

The extreme demand during the pandemic provides a new opportunity to
understand how infrastructure accessibility, quality, and reliability
affects telecommuting, particularly in light of the high volumes of
bandwidth-intensive video conferencing required in order to avoid the
face-to-face contact that could increase the spread of infection. We
seek to answer how internet service resilience contributes to or reduces
economic resilience when the latter is dependent upon the capability to
work from home. We also aim to improve our understanding of the impact
of first level digital division on telecommuting, and whether this
results in much more fundamental third level digital division than has
previously been perceived.

Furthermore, these multi-layered digital divides intersect with material
divides and the economic geography of the UK. Following the regional
economic resilience literature, which underlines the differentiated
capacity of cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}, different places have
different industrial and occupational profiles, and these affect the
aggregated potential capacity of places for telecommuting. Such profiles
are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide
\citep{martin_north_south}. Various studies have illustrated severe
inequalities between the north and the south regions of England
regarding, for example, skills and human capital, unemployment,
productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers some of the highest level
of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
digital divides associated with or shaped by the geography of the UK,
but the intersection of the digital and material divides affects the
capacity of places to overcome some of the economic effects of the
Covid-19 pandemic. Importantly, this is the first time that digital
technologies became an essential tool for economic resilience for such a
great part of the population.

@@

First we identified our methodological framework as cluster analysis,
which can be defined within the modern machine learning framework as an
unsupervised learning task, partitioning unlabelled observations into
homogeneous groups known as clusters \citep{montero2014tsclust}. The key
idea is that observations within clusters tend to be more similar than
observations between clusters. Clustering is particularly useful for
exploratory studies as it identifies structures within the data
\citep{aghabozorgi2015time}. Therefore, cluster analysis is a widely
used in geography \citep{gordon1977classification, everitt1974cluster}.
For instance, clustering methods are the basis of
\emph{geodemographics}, a research domain which aims to create small
area indicators or typologies of neighbourhoods based on diverse
variables \citep{SINGLETON2009289, harris2005geodemographics}. They have
also been employed to solve \emph{regionalisation} problems
\citep{niesterowicz2016}. Common characteristics of these studies are
the cross-sectional nature of the data they employ. Indeed, most
clustering problems in geography deal with observations that are fixed
in time. However, for this paper we are interested in internet speeds,
which vary over time and, therefore, we create clusters of local
authorities in the UK with similar temporal signatures of experienced
internet speeds.

To do so, we employ time-series clustering methods, which have been
developed to deal with clustering problems linked to, for instance,
stock or other financial data, economic, governmental or medical data as
well as machine monitoring
\citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
The main challenge -- and also the difference with cross-sectional
clustering problems -- is data dimensionality given the multiplicity of
data points for every individual object, local authorities in our case,
included in the data set. Time-series are dynamic data as the value of
the observations change as a function of time
\citep{aghabozorgi2015time}. This high dimensionality leads to (i)
computational and algorithmic challenges regarding handling these data
and building algorithms to perform clustering over long time-series, and
(ii) open questions regarding the choice of similarity measures in order
to cluster similar times-series objects together considering the whole
length of the time-series and overcoming issues around noise, outliers
and shifts \citep{lin2004iterative, aghabozorgi2015time}.

For this paper we utilise a category of time-series clustering methods
known as shape-based approaches. These methods match two sperate
time-series objects based on the similarity of their shapes through the
calculation of distances between the shapes, and are better equipped to
capture similarities between short length time-series
\citep{aghabozorgi2015time}, such as our data. We thus identify clusters
of UK local authorities with similar temporal signatures -- i.e.~shapes
-- of experienced internet speeds. The clusters are identified using a
partitioning algorithm, which leads to non-overlapping clusters, and
particularly the widely use \emph{k}-means algorithm because of the
simplicity of the implementation and the interpretability of the
results. This is an iterative algorithm, which begins with defining the
desired number of clusters \emph{k}. Then each observation is randomly
assigned to a cluster from the \([1,k]\) space. This initial cluster
assignment is followed by iterations in order to minimise the distance
between the centroids of the clusters and the observations assigned to
these clusters \citep{james2013introduction}.

@@

The frequency of speed tests was important in identifying the temporal
profile which would give us most insight into experienced internet
service and resilience over units of time. Whilst there is an overall
trend of increased testing from March to April and then a slight
reduction from April to May, this trend masks substantial variation by
not only the day of the week, but also time of day, as can be seen in
Figure \ref{test2020}. Thus, a daily aggregation of upload speeds would
mask the variation in experienced service over the course of each
weekday. Furthermore, the importance of this variation is highlighted by
a comparison with the same period in \(2019\), as in Figure
\ref{test2019}, when the volume of testing and thus of experience of
internet service quality peaked in the evening, presumably in response
to demand for leisure activities and download speeds. In contrast, the
majority of the increase in testing in \(2020\) is during the working
day, creating a new morning peak in Figure \ref{test2020}. Therefore, we
include a measure of hourly variation in our temporal profiles to
reflect the change in users' perception of the workday reliability of
internet services.

@@

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth,height=0.4\textheight]{figures/time.var.plot2020.png}
\caption{Speed tests over time, 2020 \label{test2020}}
\end{figure}

@@

The cause of these different experiences of broadband resilience may
vary across different areas, as they may reflect either similarities in
patterns of demand or similar quality of infrastructure. Our approach is
also limited by potential endogeneity, as for example, better quality
connections with high mean speeds may enable more working from home, but
greater demand may cause slower speeds, less reliability and greater
variability of speed at different times of day or week. Therefore, we
avoid attributing any cause to our analysis of the experienced level of
quality and reliability of upload speeds. Instead, we run an auxiliary
regression to understand how the spatial and temporal patterns of
internet service relate to the economic geography of the UK. More
specifically, we estimate the following multinomial logit model:

@@

Based on the outcomes of the time-series clustering, we identify \(J\)
distinct and crisp clusters. We then regress this cluster membership
against a vector \(X_{i}\) of socio-economic and geographic variables,
which are discussed in detail in the relevant Section
\protect\hyperlink{sec:4.2}{4.2}. Because we cannot identify individuals
or households and, consequently, we aggregated our data at the LAD
level, our results offer correlations between the socioeconomic
characteristics of certain geographic locations and internet service
quality, not a record of who was telecommuting. Such individual data
could be found though surveys, but these offer less detailed information
about the experience of internet resilience due to enforced demand,
which is the main contribution of this paper. Our auxiliary regression,
therefore, provides an indication of how internet connectivity can
reinforce or redress existing spatial and social inequalities in
different places. However, it opens a path to future research by
highlighting the importance of\\
understanding of how telecommuting capabilities and digital
infrastructure divisions intersect.

@@

Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
LADs and \(29.5\) million people -- experienced aggregate mean upload
speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
lower than the other three large clusters and most of the smaller
clusters, suggesting a middling quality of service. The temporal profile
for cluster \(6\) in Figure \ref{UpClusterL} shows that upload speeds
are highest at \(6:00\) on a Monday morning and peak between \(23:00\)
and midnight on Wednesday and Thursday, but tend to be lower during the
working day. Furthermore, experienced mean upload speeds in the morning
peak are \(4.4\)\% lower than in the evening peak -- a greater, more
noticeable change than any of the other large clusters experience,
suggesting poorer reliability during the working day. This difference is
less, however than any of the clusters as shown in Figure
\ref{UpClusterS}.

Clusters \(8\) and \(12\) also have mean upload speeds under \(10\)Mb/s,
but higher than clusters \(1\), \(13\), and \(6\). However, mean upload
speeds are much lower between \(9:00-10:59\) than between
\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning
than in the evening in cluster \(12\). Indeed, cluster \(12\) is the
only cluster to experience higher speeds in the morning peak compared to
the evening, and thus the only cluster where the temporal profile of
internet use is closer to what might have been expected pre-pandemic.
And yet even among the other clusters, the reliability of internet
services during the working day still varies considerably. Interpreting
this variation from the large spikes and dips shown on Figures
\ref{UpClusterL} and \ref{UpClusterS} is difficult, but the statistics
in Table \ref{up.cluster.descr} show that clusters \(9\) and \(11\) have
the most reliable internet services. The ratio of standard deviation to
mean in both these clusters is lowest, and speeds are only about 2\%
slower in the morning than the evening. Mean speeds are also higher than
in any other cluster, excluding cluster \(2\), where measures of
reliability suggests poorer performance.

Thus, broadband services in clusters \(9\) and \(11\), home to over
twelve million people performed the best during the study period, in
terms of both quality and reliability. In Figure \ref{UpClusterL},
cluster \(11\) shows more noticeable peaks and troughs, but the lowest
points are not during the morning peak, but occur, on average, between
\(6:00-7:00\) on Monday morning, \(14:00-15:00\) on Friday, and
\(16:00-17:00\) on Thursday. These slow periods still offer better
speeds than the average hourly profile of cluster \(13\). Finally,
ignoring the smallest clusters in terms of population, that is clusters
\(4\) and \(5\), clusters \(3\), \(7\) and \(10\) also have relatively
high mean upload speeds. Clusters \(3\) and \(10\) pair high mean speeds
with low standard deviations relative to the mean speeds, suggesting
reliability and resilience, as well as quality broadband services.
Cluster \(7\) has a higher ratio of standard deviation to mean, but
there is less difference in average speeds between the morning and
evening peaks than in clusters \(3\) and \(10\).

@@

Using an auxiliary multinomial regression, we test whether the clusters
that have higher mean speeds and more reliable services consist of LADs
that are more urban, affluent, and / or more likely to benefit from a
choice of high quality internet services. We also estimate which of our
clusters are more likely to have a higher proportion of occupations
which enable telecommuting because of the nature of the work. The
results of the auxiliary regression are presented in Table \ref{aux}.
The dependent variable is the LAD cluster membership as described in the
\protect\hyperlink{sec:3}{methods and data} section and equation
\ref{eq1} and each column represent a different cluster. The reference
case is cluster \(4\), which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand
people. Mean, experienced upload speeds in cluster \(4\) (see Table
\ref{up.cluster.descr}) are close to both the average speeds for the
\(13\) clusters (\(9.9\)Mb/s) and the pre-clustered average for the
whole sample (\(9.3\)Mb/s). However, the standard deviation for cluster
\(4\) and the difference between average speeds in the morning compared
to the evening peak periods are indications of worse reliability than
many of the other clusters. Hence, the results in Table \ref{aux} should
be seen as relative rather than absolute probabilities.

First, we control for the number of speed tests run per cluster
inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
Virgin Media internet connections. Regarding the former, we expect
people in LADs with more unreliable connections to test their internet
speeds more often, and the results validate our priors. Meanwhile, fast
Virgin Media cable connections have historically only been available to
\(45\)\% of premises in the UK \citep{ofcom2016}, where the more
lucrative and competitive market originally attracted the cable TV
provider. Those in clusters \(2\), \(3\), \(9\) and \(11\) benefit from
a higher proportion of Virgin connections, which is an indication that
people in these clusters are more likely to live in urban areas, with
more choice of broadband services, or in other words, are more likely to
be on the right side of the first level, infrastructure-based digital
divide, as we expected from the analysis in Section
\protect\hyperlink{sec:4.1}{4.1}.

Then, we employ distance from London and from the nearest metropolitan
area (including London) as two variables depicting peripherality, urban
structure and, potentially, first level digital divides. The broadband
speed tests run in the authorities in cluster \(3\) are more likely to
be taking place close to London than those run in any of the other
clusters, and two of the four authorities in cluster \(3\) are the
London commuter towns of Harlow and Luton. However, even though London
was also included in the variable calculating distance from the centre
of one of either the ten largest metropolitan areas in England, or
Glasgow or Cardiff, tests run in cluster \(3\) are likely to be furthest
away. Thus, it is important to consider the membership of each cluster
as well as the regression results. Corby and Eastbourne, the other two
authorities in cluster \(3\) are, whilst not part of metropolitan areas,
are also large, accessible towns.

@@

Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
of a large metropolitan area, even though the four local authorities of
cluster \(7\) include no central urban boroughs and only one LAD that is
part of a metropolitan area of governance - Tameside in Greater
Manchester. This may explain why those in cluster \(7\) are the second
least likely to be served by Virgin Media. It is also a demonstration of
the complexity of experienced broadband upload speeds as captured by
time-series clustering, and how quality and reliability, not only
availability, engender first level digital divides. Cluster \(1\), which
our analysis suggests lacks broadband resilience, contains five, mainly
rural authorities. However, they are scattered around the country --
Ceredigion in West Wales, Darlington in the Northeast, Eden and
Rossendale in the Northwest, and Rutland in the East Midlands -- and
therefore, the results in Table \ref{aux} indicate that these
authorities are closer to a metropolitan area than authorities in
cluster \(3\), but furthest from London.

>>>>

\begin{abstract}
This paper offers a new perspective on telecommuting from the viewpoint
of the complex web of digital divides. Using the UK as a case study,
this paper studies how the quality and reliability of internet services,
as reflected in \emph{experienced} internet upload speeds during the
spring 2020 lockdown, may reinforce or redress the spatial and social
dimensions of digital divisions. Fast, reliable internet connections are
necessary for the population to be able to work from home. Although not
every place hosts individuals in occupations which allow for
telecommuting nor with the necessary skills to effectively use the
internet to telecommute, good internet connectivity is also essential to
local economic resilience in a period like the current pandemic.
Employing data on individual broadband speed tests and state-of-the-art
time-series clustering methods, we create clusters of UK local
authorities with similar temporal signatures of experienced upload
speeds. We then associate these clusters of local authorities with their
socioeconomic and geographic characteristics to explore how they overlap
with or diverge from the existing economic and digital geography of the
UK. Our analysis enables us to better understand how the spatial and
social distribution of both occupations and online accessibility
intersect to enable or hinder the practice of telecommuting at a time of
extreme demand.
\end{abstract}

@@

In this analysis, the terms `telecommuting' and `working from
home\footnote{See also the popular social media hashtag \#WFM}' are used
interchangeably, as most remote labour during the Covid-19 crisis was
carried out in the homes of individual employees rather than any other
location \citep{eurofound2020}. However, previous research has explored
how telecommuting can occur in other places, including satellite offices
or on public transport \citep{felstead2012rapid, siha2006telecommuting}.
Previous research has also used a variety of definitions to measure the
level of telecommuting within different workforces, distinguishing
between those directly employed, indirectly employed, self-employed,
full-time or part-time, and those who use digital technologies to work
remotely full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. Studies from the United States, the Netherlands, and
the UK found that telecommuters are most likely to hold professional,
managerial, and technical occupations where the workforce is better
educated and wealthier, and that there is suppressed demand among women
and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.

Opportunities for working from home during the current pandemic have
likewise not been equally spread across the workforce.
\citet{NBERw26948} indicated that in the US, managers, educators, those
working in computer-related occupations, finance, and law can easily
work from home, and that occupations with opportunities to telecommute
are associated with higher earnings. This is not the case for the
workforce occupied in more spatially fixed occupations, from farming,
construction and manufacturing to hospitality and care services. In the
US, these occupations tend to be lower-income, non-white, without a
university degree, live in rental accommodation and lack health
insurance \citep{NBERw27085}. Similar trends can be observed for other
countries. For example, \(75\)\% of workers with tertiary education
worked from home in Europe during spring \(2020\), whilst only \(34\)\%
of workers with secondary education and \(14\)\% of those primary
education did so \citep{eurofound2020}.

@@

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
These returns have been even greater during the Covid-19 crisis, when
those with the capability to telecommute also have the ability to
maintain their employment whilst protecting their health. However, the
success of these arrangements has been dependent upon the first level
digital divide, which is associated with access to and quality of
internet connectivity. \citet{SALEMINK2017360} provides a systematic
review of the pre-pandemic, first level digital divide in infrastructure
quality between urban and rural areas in various advanced economies.
Rural areas, predictably, fare worse. Yet whether this variation in
infrastructure quality affects the spatial footprint of telecommuting
has not previously been investigated, in part because telecommuting has
not previously been a cause of concentrated demand and pressure on
internet services.

There are indications that those who purchase high speed connections
consume more data of all sorts and use their connections for a greater
variety of purposes \citep{hauge2011consumer}. There is also a
correlation between access to internet services and a reduction in
household transport spend \citep{bris2017ict}. Independently of whether
the additional internet use and reduced travel is because of increased
telecommuting, these studies suggest that better internet services
enable households to make savings and efficiencies, an example of the
first level digital divide reinforcing the third level.

Multi-layered digital divides may also intersect with material divides
and the economic geography of the UK. The regional economic resilience
literature underlines the differentiated capacity of cities and regions
to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}. As different places have
different industrial and occupational profiles, these affect their
aggregated potential capacity for telecommuting. Such profiles are
associated with longstanding inequalities in the UK and their spatial
representation as a North-South divide \citep{martin_north_south}.
Various studies have illustrated severe inequalities between the north
and the south regions of England in terms of skills and human capital,
unemployment, productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers from some of the highest
levels of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. All three levels of digital
divides are associated with or shaped by the geography of the UK. Yet
this is the first time that the intersection of digital and material
divides is relevant to understanding the economic resilience of places
and large swathes of the population, as digital technologies became an
essential tool of productivity during the Covid-19 pandemic.

The extreme demand during the pandemic thus provides a new opportunity
to understand how internet infrastructure quality, and reliability
affects telecommuting, particularly in light of the high volumes of
bandwidth-intensive video conferencing required in order to avoid the
face-to-face contact that could increase the spread of infection. We
seek to answer how internet service resilience contributes to or reduces
economic resilience when the latter is dependent upon the capability to
work from home. We also aim to improve our understanding of the impact
of first level digital division on telecommuting, and whether this
results in more fundamental third level digital divisions than has
previously been perceived.

@@

Our chosen methodological framework is cluster analysis, which can be
defined within machine learning approaches as an unsupervised learning
task, partitioning unlabelled observations into homogeneous groups known
as clusters \citep{montero2014tsclust}. The key idea is that
observations within clusters tend to be more similar than observations
between clusters. Clustering is particularly useful for exploratory
studies as it identifies structures within the data
\citep{aghabozorgi2015time}. Cluster analysis is a widely used in
geography \citep{gordon1977classification, everitt1974cluster}, for
example to solve \emph{regionalisation} problems
\citep{niesterowicz2016}. Clustering methods are also the basis of
\emph{geodemographics}, a research domain which aims to create small
area indicators or typologies of neighbourhoods based on diverse
variables \citep{SINGLETON2009289, harris2005geodemographics}. These
studies usually employ cross-sectional data, and most clustering
problems in geography deal with observations that are fixed in time.
However, for this paper we are interested in internet speeds, which vary
over time. Therefore, we create clusters of local authorities in the UK
with similar temporal signatures of experienced internet speeds.

To do so, we employ time-series clustering methods, which have been
developed to deal with clustering problems linked to, for instance,
stock or other financial data, economic, governmental or medical data as
well as machine monitoring
\citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
The main challenge, which does not apply to cross-sectional clustering
problems, is data dimensionality, with a multiplicity of data points for
every individual object included in the data set, and how their value
changes dynamically as a function of time \citep{aghabozorgi2015time}.
This high dimensionality leads to (i) computational and algorithmic
challenges regarding handling these data and building algorithms to
perform clustering over long time-series, and (ii) open questions
regarding the choice of similarity measures in order to cluster similar
times-series objects together considering the whole length of the
time-series and overcoming issues around noise, outliers and shifts
\citep{lin2004iterative, aghabozorgi2015time}.

For this paper we utilise a category of time-series clustering methods
known as shape-based approaches. These methods match two separate
time-series objects based on the similarity of their shapes through the
calculation of distances between the shapes, and are better equipped to
capture similarities between short length time-series
\citep{aghabozorgi2015time}, such as our data. We thus identify clusters
of UK local authorities with similar temporal signatures -- i.e.~shapes
-- of experienced internet speeds. The clusters are identified using the
common partitioning algorithm, where no clusters overlap, known as
\emph{k}-means. This iterative algorithm is popular because of the
simplicity of the implementation and the interpretability of the
results. It begins with defining the desired number of clusters:
\emph{k}. Then each observation is randomly assigned to a cluster from
the \([1,k]\) space. This initial cluster assignment is followed by
iterations in order to minimise the distance between the centroids of
the clusters and the observations assigned to these clusters
\citep{james2013introduction}.

@@

The frequency of speed tests was important in identifying the temporal
profile which would give us most insight into experienced internet
service and resilience over units of time. Whilst there is an overall
trend of increased testing from March to April and then a slight
reduction from April to May, this trend masks substantial variation by
not only the day of the week, but also time of day, as can be seen in
Figure \ref{test2020}. Thus, a daily aggregation of upload speeds would
mask the variation in experienced service over the course of each
weekday. Furthermore, the importance of this variation is highlighted by
a comparison with the same period in \(2019\), as in Figure
\ref{test2019}, when the volume of testing and thus of experienced
internet service quality peaked in the evening, presumably in response
to demand for leisure activities and download speeds. In contrast, the
majority of the increase in testing in \(2020\) is during the working
day, creating a new morning peak in Figure \ref{test2020}. Therefore, we
include a measure of hourly variation in our temporal profiles to
reflect the change in users' perception of the workday reliability of
internet services.

@@

\begin{figure}
\centering
\includegraphics[width=0.85\textwidth,height=0.5\textheight]{figures/time.var.plot2020.png}
\caption{Speed tests over time, 2020 \label{test2020}}
\end{figure}

@@

The cause of these different experiences of broadband resilience may
vary between and within clusters, as they may reflect either patterns of
demand or quality of infrastructure. Our approach is also limited by
potential endogeneity, as for example, better quality connections with
high mean speeds may enable more working from home, but greater demand
can cause slower speeds, less reliability, or greater variability of
speed at different times of day or week. Therefore, we avoid attributing
any cause to our analysis of the experienced level of quality and
reliability of upload speeds. Instead, we run an auxiliary regression to
understand how the spatial and temporal patterns of internet service
relate to the economic geography of the UK. More specifically, we
estimate the following multinomial logit model:

@@

Based on the outcomes of the time-series clustering, we identify \(J\)
distinct and crisp clusters. We then regress this cluster membership
against a vector \(X_{i}\) of socio-economic and geographic variables,
which are discussed in detail in the relevant Section
\protect\hyperlink{sec:4.2}{4.2}. Because we cannot identify individuals
or households and consequently aggregated our data at the LAD level, our
results offer correlations between the socioeconomic characteristics of
certain geographic locations and internet service quality, not a record
of who was telecommuting. Such individual data could be found though
surveys, but these offer less detailed information about the experience
of internet resilience due to enforced demand, which is the main
contribution of this paper. Our auxiliary regression, therefore,
provides an indication of how internet connectivity can reinforce or
redress existing spatial and social inequalities in different places.
However, it opens a path to future research by highlighting the
importance of understanding of how telecommuting capabilities and
digital infrastructure divisions intersect.

@@

Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
LADs and \(29.5\) million people -- experienced aggregate mean upload
speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
lower than the other three large clusters and most of the smaller
clusters, suggesting a middling quality of service. The temporal profile
for cluster \(6\) in Figure \ref{UpClusterL} shows that upload speeds
are highest at \(6:00\) on a Monday morning and between \(23:00\) and
midnight on Wednesday and Thursday, but tend to be lower during the
working day. Furthermore, experienced mean upload speeds in the morning
peak are \(4.4\)\% lower than in the evening peak -- a greater, more
noticeable change than any of the other large clusters experience,
suggesting poorer reliability during the working day. This difference is
less, however than any of the clusters included in Figure
\ref{UpClusterS}.

Clusters \(8\) and \(12\) also have mean upload speeds under \(10\)Mb/s,
but higher than clusters \(1\), \(13\), and \(6\). However, mean upload
speeds are much lower between \(9:00-10:59\) than between
\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning
than in the evening in cluster \(12\). Indeed, cluster \(12\) is the
only cluster to experience higher speeds in the morning peak compared to
the evening, and thus the only cluster where the temporal profile of
internet use is closer to what might have been expected pre-pandemic.
And yet even among the other clusters, the reliability of internet
services during the working day still varies considerably. Interpreting
this variation from the large spikes and dips shown on Figures
\ref{UpClusterL} and \ref{UpClusterS} is difficult, but the statistics
in Table \ref{up.cluster.descr} show that clusters \(9\) and \(11\) have
the most reliable internet services. The ratio of standard deviation to
mean in both these clusters is lowest, and speeds are only about 2\%
slower in the morning than the evening. Mean speeds are also higher than
in any other cluster, excluding cluster \(2\), where measures of
reliability suggest poorer performance.

Thus, broadband services in clusters \(9\) and \(11\), home to over
twelve million people, performed the best during the study period, in
terms of both quality and reliability. In Figure \ref{UpClusterL},
cluster \(11\) shows more noticeable peaks and troughs, with the lowest
points occurring, on average, between \(6:00-7:00\) on Monday morning,
\(14:00-15:00\) on Friday, and \(16:00-17:00\) on Thursday, not the
workday peak. Furthermore, these slow periods offer better speeds than
the average hourly profile of cluster \(13\). Finally, ignoring the
smallest clusters in terms of population, that is clusters \(4\) and
\(5\), clusters \(3\), \(7\) and \(10\) also have relatively high mean
upload speeds. Clusters \(3\) and \(10\) pair high mean speeds with low
standard deviations relative to the mean speeds, suggesting reliability
and resilience, as well as quality broadband services. Cluster \(7\) has
a higher ratio of standard deviation to mean, but there is less
difference in average speeds between the morning and evening peaks than
in clusters \(3\) and \(10\).

@@

Using an auxiliary multinomial logit regression, we test whether the
clusters that have higher mean speeds and more reliable services consist
of LADs that are more urban, affluent, and / or more likely to benefit
from a choice of high quality internet services. We also estimate which
of our clusters are more likely to have a higher proportion of
occupations where the nature of the work enables telecommuting. The
results of the auxiliary regression are presented in Table \ref{aux}.
The dependent variable is the LAD cluster membership as described in the
\protect\hyperlink{sec:3}{methods and data} section and equation
\ref{eq1}. Each column represents a different cluster. The reference
case is cluster \(4\), which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand
people. Mean, experienced upload speeds in cluster \(4\) (see Table
\ref{up.cluster.descr}) are between the average speeds for the \(13\)
clusters (\(9.9\)Mb/s) and the pre-clustered average for the whole
sample (\(9.3\)Mb/s). However, the standard deviation for cluster \(4\)
and the difference between average speeds in the morning compared to the
evening peak periods are indications of worse reliability than many of
the other clusters. Hence, the results in Table \ref{aux} should be seen
as relative rather than absolute probabilities.

First, we control for the number of speed tests run per cluster
inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
Virgin Media internet connections. Regarding the former, we expect
people in LADs with more unreliable connections to test their internet
speeds more often, and the results validate our priors. Meanwhile, fast
Virgin Media cable connections have historically only been available to
\(45\)\% of premises in the UK \citep{ofcom2016}, where the more
lucrative and competitive market originally attracted the cable TV
provider. Those in clusters \(2\), \(3\), \(9\) and \(11\) benefit from
a higher proportion of Virgin connections, which is an indication that
people in these clusters are more likely to live in urban areas, with
more choice of broadband services. In other words, they are more likely
to be on the right side of the first level, infrastructure-based digital
divide, as we expected from the analysis in Section
\protect\hyperlink{sec:4.1}{4.1}.

We employ distance from London and from the nearest metropolitan area
(including London) as two variables depicting peripherality, urban
structure and, potentially, first level digital divides. The broadband
speed tests run in the authorities in cluster \(3\) are more likely to
be taking place close to London than those run in any of the other
clusters, and two of the four authorities in cluster \(3\) are the
London commuter towns of Harlow and Luton. However, even though London
was also included in the variable calculating distance from the centre
of one of either the ten largest metropolitan areas in England, or
Glasgow or Cardiff, tests run in cluster \(3\) are likely to be furthest
away. Thus, it is important to consider the membership of each cluster
as well as the regression results. Corby and Eastbourne, the other two
authorities in cluster \(3\) are large, accessible towns, although not
part of metropolitan areas.

@@

Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
of a large metropolitan area, even though the four local authorities of
cluster \(7\) include no central urban boroughs and only one LAD that is
part of a metropolitan area of governance - Tameside in Greater
Manchester. This may explain why those in cluster \(7\) are the second
least likely to be served by Virgin Media. It is also a demonstration of
the complexity of both experienced broadband upload speeds as captured
by time-series clustering, and the geography of first level digital
divides as a product of quality as well as availability. In comparison,
cluster \(1\), which our analysis suggests lacks broadband resilience,
contains five, mainly rural authorities, but they are closer to a
metropolitan area than authorities in cluster \(3\), although furthest
from London, perhaps because they are scattered around the country.

Internet resilience is also more nuanced than our -- arguably crude --
dummy variable depicting the North-South economic divide, which assigns
\(1\) to LADs located in Greater London, Southeast, Southwest and East
of England regions, and \(0\) to the rest. The authorities most likely
to be in the South are those in cluster \(13\), which was identified as
having the slowest mean upload speeds of any of the clusters, and thus a
low level of service. However, cluster \(13\) includes some rural,
remote areas of the country, such as Northwest Scotland, Cornwall and
Powys in Wales as shown on Figure \ref{map.up.clusters}. It also
includes major metropolitan centres in the North -- Liverpool, Newcastle
-- and South -- Bristol, and nine (of \(32\)) London Boroughs. There are
also plenty of Southern home county and suburban areas. The standard
deviation measure for cluster \(13\) is high, but speed variation is low
during the morning peak, suggesting that the estimates for reliability
are inconsistent. Considering that this is one of the largest clusters,
and thus the averages incorporate more noise than some of the smaller
clusters, it may be that the LADs in this cluster do not all suffer
equally from a first level digital divide.
